Comparative Analysis of Parallel Computational Schemes for Solving the Time-consuming Decision Making Problems

### 2019-11-19 2019/PACT_MCO/source/2019_pavt.tex
Добавил исходники статей по ГО за 2019г. (Гергель Козинов)

In the present paper, the results of the performed investigations on the generalization of the decision making problem statements \cite{x11,x12} and on the development of the highly efficient global optimization methods utilizing the whole search information obtained in the course of computations \cite{x13,x14,x15,x16,x17} are presented.

@@

The values of characteristics $w(y)$ are supposed to be non-negative, and the decreasing of these ones corresponds to the increasing of the efficiency of the chosen variants. Also, it is supposed that the characteristics $w_j (y)$, $1 \leq j \leq M$ can be multiextremal and the computing of the values of these ones can require a large amount of computations. Besides, within the framework of considered approach, the characteristics $w_j(y)$, $1 \leq j \leq M$ are supposed to satisfy the Lipschitz condition 
\begin{equation}
\label{eq:03}
|w_j(y_1) - w_j(y_2)| \leq L_j \|y_1-y_2\|, 1 \leq j \leq M,
\end{equation}
where $L_j$ is the Lipschitz constant for the characteristic $w_j(y)$, $1 \leq j \leq M$, and  $\|*\|$ denotes the Euclidean norm in $R^N$. It is important to note that the fulfillment of the Lipschitz condition corresponds to the practical applications -- at small variations of the parameters $y \in D$ the corresponding variations of the values of characteristics $w_j(y)$, $1 \leq j \leq M$ are limited as a rule.

@@

2. The requirements to the optimality of the chosen variants $y \in D$ of the DMP can be defined in the following way. First of all, the characteristics $w_j(y)$, $1 \leq j \leq M$ should be selected, for which the achievement of the minimum possible values is necessary. Defining a set of indices $F=\{ i_1,i_2, \dots ,i_s \}$ of such characteristics defines the vector criterion of efficiency 
\begin{equation}
\label{eq:05}
f(y)=(w_{i_1}(y),w_{i_2}(y),\dots,w_{i_s}(y)).
\end{equation}

For the characteristics $w_j(y)$, $1 \leq j \leq M$ not included into the vector criterion of efficiency with the set of indices $G=\{ j_1,j_2, \dots ,j_m\}$ the sufficient levels of efficiency should be defined by a tolerance vector $q=(q_1,q_2,\dots,q_m)$. The availability of tolerances allows defining a vector function of constraints 
\begin{equation}
\label{eq:06}
g(y)= (g_1(y),g_2(y),\dots,g_m (y)), g_l (y) = w_{j_l}(y) - q_l, j_l \in G, 1 \leq l \leq m.
\end{equation}

@@

Due to a high relevance, a large enough quantity of methods of solving the MCO problems have been developed and used widely -- see, for example, \cite{x2,x3,x4,x5,x6,x7,x8,x9,x10}. A part of these algorithms ensures the obtaining the numerical estimates of the whole Pareto set $PD(Q)$ \cite{x3,x26,x27,x28}. Along with the usefulness of such an approach (all effective decisions of the MCO problem are found), the use of these methods appears to be difficult often because of high computational complexity of obtaining the estimates of the Pareto set. Besides, the estimate of the whole set $PD(Q)$ may appear tо be redundant when to solve the MCO problem it is enough to obtain several particular effective decisions (that takes place in the practical applications often). As a result, more widely used approach to solving the MCO problems is based on the scalarization of the vector criterion into some general scalar criterion of efficiency, the optimization of which can be performed with the use of already existing optimization methods. Among these methods, for example, there are weighted sum method, compromise programming method, weighted min-max method, goal programming, and many other algorithms -- see, for example, \cite{x2,x3,x4,x5,x6}.

The general approach used in the present work consists in the reduction of solving a MMGO problem to the solving of a series of single-criterion global optimization problems \footnote{Due to initial assumptions on possible multiextremality of the characteristics $w_j(y), 1 \leq j \leq M$ from (\ref{eq:01})}: 
\begin{equation}
\label{eq:12}
\min {\varphi(y)}=F(\lambda,y), g(y) \leq 0, y \in D, 
\end{equation}
where $g(y)$ is the vector function of constraints from (\ref{eq:06}), and $F(\lambda,y)$ is the minimized convolution of the partial criteria $f_i(y(x))$, $1 \leq i \leq s$ of the MCO problem (\ref{eq:05})
\begin{equation}
\label{eq:13}
F(\lambda,y)=\max{⁡(\lambda_i f_i (y), 1 \leq i \leq s)} 
\end{equation}
using the vector of the weighting coefficients 
\begin{equation}
\label{eq:14}
\lambda=(\lambda_1, \lambda_2, \dots , \lambda_s )\in \Lambda \subset R^s: \sum_{i=1}^s{\lambda_i=1}, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

The numerical solving of the global optimization problems is reduced usually to a sequential computing the values of characteristics $w(y)$ at the points $y^i$, $0\leq i \leq k$ of the search domain $D$ \cite{x18,x22,x23,x24,x25}. The data obtained as a result of computations can be represented in a form of matrix of search information (MSI)
\begin{equation}
\label{eq:18}
\Omega_k=\{(y^i,w^i=w(y^i ) )^T:1 \leq i \leq k\}.
\end{equation}

As a result of scalarization of the vector criterion (\ref{eq:12}) and application of the dimensionality reduction (\ref{eq:15}), $\Omega_k$ from (\ref{eq:18}) can be transformed into the form of the matrix of the search state (MSS):
\begin{equation}
\label{eq:19}
A_k=\{(x_i,z_i,g_i,l_i )^T:1 \leq i \leq k\},
\end{equation}
where
\begin{itemize}
	\item $x_i$, $1 \leq i \leq k$ are the reduced points of the executed global search iterations where the criteria values have been computed; 
	\item $z_i$, $g_i$, $1 \leq i \leq k$ are the values of the scalar criterion and of the constraints at the points $x_i$, $1 \leq i \leq k$ for current optimization problem $F(\lambda,y(x))$ from (\ref{eq:15}) being solved i. e.
\begin{equation}
z_i = \varphi(x_i) =F(\lambda,y(x_i)),  g_i=g(y(x_i )),  1\leq i \leq k;
\end{equation}
	\item $l_i$, $1 \leq i \leq k$ are the indices of the global search iterations, at which the points $x_i$, $1 \leq i \leq k$ have been generated, these indices are used to storing the correspondence between the reduced points and the multidimensional ones at the executed iterations i. e.
\begin{equation}
y^j = y(x_i), j = l_i, 1 \leq i \leq k.
\end{equation}

@@

The matrix of the search state $A_k$ contains the search information transformed to current scalar reduced problem (\ref{eq:15}) being solved. Besides, the search information in the MSS is arranged according to the values of coordinates of the points $x_i$, $1 \leq i \leq k$ for more efficient execution of the global search algorithms; the arranged representation of the points is reflected by the use of the lower index i. e.
\begin{equation}
x_1 \leq x_2 \leq \dots \leq x_k.
\end{equation}

The representation of the search information obtained in the course of computations in the form of  the matrices $\Omega_k$ and $A_k$ makes the basis for the development of efficient optimization search procedures. The availability of such information allows performing an adaptive choice of the points for executing the global search iterations taking into account the results of all computations completed earlier 
\begin{equation}
\label{eq:20}
y^{k+1}=Y(\Omega_k),k=1,2,\dots.
\end{equation}
where $Y$ is the rule for computing the points $y^i$, $0 \leq i \leq k$ of the applied optimization algorithm. Such an adaptive choice of the points of executed search iterations can accelerate the finding of the effective decisions. In the case of the global optimization problems, the accumulation of the whole search information and the use of the rules of the kind (\ref{eq:20}) is just mandatory. Any reduction of the search information in determining the points to be executed would result in executing the excess global search iterations.

It should be noted that the availability of MSS $\Omega_k$ from (\ref{eq:18}) allows reducing the results of all preceding computations $z_i$,  $1 \leq i \leq k$ stored in the MSS $A_k$ to the values of current optimization problem $F(\lambda,y(x))$ from (\ref{eq:15}) being solved for any new parameter of the MCO problem statement $P$ from (\ref{eq:09}) and for any new value of the convolution coefficients $\lambda$ from (\ref{eq:14}) on the basis of the values of characteristics $w^i$, $1 \leq i \leq k$ computed earlier and stored in $\Omega_k$ without any repeated time-consuming computation of the values of $w(y)$ from (\ref{eq:01}) i.e.
\begin{equation}
\label{eq:21}
w_i \xrightarrow{\lambda,P} (z_i,g_i ), 1 \leq i \leq k,\forall \lambda \in \Lambda, P \in \mathbb{P}.
\end{equation}

@@

The method of organizing the parallel computations considered above can be developed further by utilizing the information compatibility of the problems from the set $\mathbb{F}$ -- as has been noted in Section {\ref{sec:04}}, the values of characteristics $w^i$, $1 \leq i \leq k$ computed when solving any problem $F_1(\lambda,y(x))$ can be transformed to the values of any other problem $F_2(\lambda,y(x))$. This result allows proposing the following general scheme of parallel computations for simultaneous solving the problems from the set $\mathbb{F}$.

@@

\textbf{3. The execution of the global search iterations.} The execution of the global search iterations for each problem $F(\lambda,y(x))$ from the set $\mathbb{F}$ being solved is performed in parallel. The execution of every iteration on each computational node includes performing the following operations:
\begin{enumerate}
	\item The choice of $q$, $q \geq 1$ points $x^i$, $1\leq i\leq q$ of the next global search iteration is performed according to the rule (\ref{eq:20}) of the optimization method; the choice of the points is performed taking into account the available search information $\Omega_k$ and $A_k$; the number of generated points $q$, $q\geq 1$ is determined by the number of computational cores employed for solving the problem $F(\lambda,y(x))$;
	\item For each chosen one-dimensional point $x^i \in [0,1]$, $1 \leq i \leq q$ of current global search iteration, the multidimensional image $y^i \in D$, $1 \leq i \leq q$ is determined according to the mapping $y(x)$. Then, each computed image $y^i \in D$, $1 \leq i \leq q$ is transmitted to all employed computational nodes to avoid the multiple choice of the same points in the domain $D$ when solving the problems from the set $\mathbb{F}$;
	\item The values of characteristics $w(y)$ from (\ref{eq:01}) are computed at all points $y^i \in D$, $1 \leq i \leq q$; for each particular point $y^i \in D$, $1 \leq i \leq q$, these computations are performed in parallel using different computational cores; the computed values of characteristics $w(y)$ are transmitted to all employed computational nodes to include the obtained data into the search information $\Omega_k$ and $A_k$.
\end{enumerate}

@@

According to the presented computational scheme of parallel computations, each computational unit would contain the identical copies of the matrix of search information $\Omega_k$ from (\ref{eq:18}); the matrices $A_k$ from (\ref{eq:19}) would contain different sets of the global search points $x_i$, $1\leq i\leq k$ because of the use of different evolvents $y(x)$ for the dimensionality reduction and of different values of the scalar criterion and constraints $z_i$, $g_i$, $1\leq i\leq k$ corresponding to the problems $F(\lambda,y(x))$ from the set $\mathbb{F}$ being solved.

@@

The developed general computational scheme includes the concurrent parallel computational methods for the computational systems with the shared memory, the distributed parallel computations using the multiple mappings for the dimensionality reduction, and the multilevel nesting of the parallel computations for the high-performance computational systems. In general, the proposed computational scheme ensures the opportunity of an efficient application of the high-performance computational systems with a large number (tens and hundreds thousand) cores/processors for solving the complex global optimization problems. Besides, this general scheme can be used for the organization of the parallel computations for a wide range of algorithms for solving the time-consuming problems of decision making (in particular, for the information-statistical global optimization algorithms).

@@

>>>>

In the present paper, the results of the performed investigations on the generalization of the decision making problem statements \cite{x11,x12} and on the development of the highly efficient global optimization methods utilizing the whole search information obtained in the course of computations \cite{x13,x14,x15,x16,x17} are presented. Parallel algorithms developed earlier are presented in papers \cite{x14,x15,x16,x17}. The proposed algorithms are able to efficiently use hundreds and thousands of computational cores \cite{x17_1}. In the scope of this paper, a comparative analysis of the implemented parallel algorithms on modern computing systems is carried out. The implementations of algorithms for shared memory, distributed memory and heterogeneous computing systems are compared. As a result of a comparative analysis, an optimal running configuration is selected.

@@

The values of characteristics $w(y)$ are supposed to be non-negative, and the decreasing of these ones corresponds to the increasing of the efficiency of the chosen variants. Also, it is supposed that the characteristics $w_j (y)$, $1 \leq j \leq M$, can be multiextremal and the computing of the values of these ones can require a large amount of computations. Besides, within the framework of considered approach, the characteristics $w_j(y)$, $1 \leq j \leq M$, are supposed to satisfy the Lipschitz condition 
\begin{equation}
\label{eq:03}
|w_j(y_1) - w_j(y_2)| \leq L_j \|y_1-y_2\|, 1 \leq j \leq M,
\end{equation}
where $L_j$ is the Lipschitz constant for the characteristic $w_j(y)$, $1 \leq j \leq M$, and  $\|*\|$ denotes the Euclidean norm in $R^N$. It is important to note that the fulfillment of the Lipschitz condition corresponds to the practical applications -- at small variations of the parameters $y \in D$ the corresponding variations of the values of characteristics $w_j(y)$, $1 \leq j \leq M$, are limited as a rule.

@@

2. The requirements to the optimality of the chosen variants $y \in D$ of the DMP can be defined in the following way. First of all, the characteristics $w_j(y)$, $1 \leq j \leq M$, should be selected, for which the achievement of the minimum possible values is necessary. Defining a set of indices $F=\{ i_1,i_2, \dots ,i_s \}$ of such characteristics defines the vector criterion of efficiency 
\begin{equation}
\label{eq:05}
f(y)=(w_{i_1}(y),w_{i_2}(y),\dots,w_{i_s}(y)).
\end{equation}

For the characteristics $w_j(y)$, $1 \leq j \leq M$, not included into the vector criterion of efficiency with the set of indices $G=\{ j_1,j_2, \dots ,j_m\}$ the sufficient levels of efficiency should be defined by a tolerance vector $q=(q_1,q_2,\dots,q_m)$. The availability of tolerances allows defining a vector function of constraints 
\begin{equation}
\label{eq:06}
g(y)= (g_1(y),g_2(y),\dots,g_m (y)), g_l (y) = w_{j_l}(y) - q_l, j_l \in G, 1 \leq l \leq m.
\end{equation}

@@

Due to a high relevance, a large enough quantity of methods of solving the MCO problems have been developed and used widely -- see, for example, \cite{x2,x3,x4,x5,x6,x7,x8,x9,x10}. A part of these algorithms ensures the obtaining the numerical estimates of the whole Pareto set $PD(Q)$ \cite{x3,x26,x27,x28}. Along with the usefulness of such an approach (all effective decisions of the MCO problem are found), the use of these methods appears to be difficult often because of high computational complexity of obtaining the estimates of the Pareto set. Besides, the estimate of the whole set $PD(Q)$ may appear to be redundant when to solve the MCO problem it is enough to obtain several particular effective decisions (that takes place in the practical applications often). As a result, more widely used approach to solving the MCO problems is based on the scalarization of the vector criterion into some general scalar criterion of efficiency, the optimization of which can be performed with the use of already existing optimization methods. Among these methods, for example, there are weighted sum method, compromise programming method, weighted min-max method, goal programming, and many other algorithms -- see, for example, \cite{x2,x3,x4,x5,x6}.

The general approach used in the present work consists in the reduction of solving a MMGO problem to the solving of a series of single-criterion global optimization problems \footnote{Due to initial assumptions on possible multiextremality of the characteristics $w_j(y), 1 \leq j \leq M$, from (\ref{eq:01})}: 
\begin{equation}
\label{eq:12}
\min {\varphi(y)}=F(\lambda,y), g(y) \leq 0, y \in D, 
\end{equation}
where $g(y)$ is the vector function of constraints from (\ref{eq:06}), and $F(\lambda,y)$ is the minimized convolution of the partial criteria $f_i(y(x))$, $1 \leq i \leq s$, of the MCO problem (\ref{eq:05})
\begin{equation}
\label{eq:13}
F(\lambda,y)=\max{⁡(\lambda_i f_i (y), 1 \leq i \leq s)} 
\end{equation}
using the vector of the weighting coefficients 
\begin{equation}
\label{eq:14}
\lambda=(\lambda_1, \lambda_2, \dots , \lambda_s )\in \Lambda \subset R^s: \sum_{i=1}^s{\lambda_i=1}, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

The numerical solving of the global optimization problems is reduced usually to a sequential computing the values of characteristics $w(y)$ at the points $y^i$, $0\leq i \leq k$, of the search domain $D$ \cite{x18,x22,x23,x24,x25}. The data obtained as a result of computations can be represented in a form of matrix of search information (MSI)
\begin{equation}
\label{eq:18}
\Omega_k=\{(y^i,w^i=w(y^i ) )^T:1 \leq i \leq k\}.
\end{equation}

As a result of scalarization of the vector criterion (\ref{eq:12}) and application of the dimensionality reduction (\ref{eq:15}), $\Omega_k$ from (\ref{eq:18}) can be transformed into the form of the matrix of the search state (MSS):
\begin{equation}
\label{eq:19}
A_k=\{(x_i,z_i,g_i,l_i )^T:1 \leq i \leq k\},
\end{equation}
where
\begin{itemize}
	\item $x_i$, $1 \leq i \leq k$, are the reduced points of the executed global search iterations where the criteria values have been computed; 
	\item $z_i$, $g_i$, $1 \leq i \leq k$, are the values of the scalar criterion and of the constraints at the points $x_i$, $1 \leq i \leq k$, for current optimization problem $F(\lambda,y(x))$ from (\ref{eq:15}) being solved i. e.
\begin{equation}
z_i = \varphi(x_i) =F(\lambda,y(x_i)),  g_i=g(y(x_i )),  1\leq i \leq k;
\end{equation}
	\item $l_i$, $1 \leq i \leq k$, are the indices of the global search iterations, at which the points $x_i$, $1 \leq i \leq k$, have been generated, these indices are used to storing the correspondence between the reduced points and the multidimensional ones at the executed iterations i. e.
\begin{equation}
y^j = y(x_i), j = l_i, 1 \leq i \leq k.
\end{equation}

@@

The matrix of the search state $A_k$ contains the search information transformed to current scalar reduced problem (\ref{eq:15}) being solved. Besides, the search information in the MSS is arranged according to the values of coordinates of the points $x_i$, $1 \leq i \leq k$, for more efficient execution of the global search algorithms; the arranged representation of the points is reflected by the use of the lower index i. e.
\begin{equation}
x_1 \leq x_2 \leq \dots \leq x_k.
\end{equation}

The representation of the search information obtained in the course of computations in the form of  the matrices $\Omega_k$ and $A_k$ makes the basis for the development of efficient optimization search procedures. The availability of such information allows performing an adaptive choice of the points for executing the global search iterations taking into account the results of all computations completed earlier 
\begin{equation}
\label{eq:20}
y^{k+1}=Y(\Omega_k),k=1,2,\dots
\end{equation}
where $Y$ is the rule for computing the points $y^i$, $0 \leq i \leq k$, of the applied optimization algorithm. Such an adaptive choice of the points of executed search iterations can accelerate the finding of the effective decisions. In the case of the global optimization problems, the accumulation of the whole search information and the use of the rules of the kind (\ref{eq:20}) is just mandatory. Any reduction of the search information in determining the points to be executed would result in executing the excess global search iterations.

It should be noted that the availability of MSS $\Omega_k$ from (\ref{eq:18}) allows reducing the results of all preceding computations $z_i$,  $1 \leq i \leq k$, stored in the MSS $A_k$ to the values of current optimization problem $F(\lambda,y(x))$ from (\ref{eq:15}) being solved for any new parameter of the MCO problem statement $P$ from (\ref{eq:09}) and for any new value of the convolution coefficients $\lambda$ from (\ref{eq:14}) on the basis of the values of characteristics $w^i$, $1 \leq i \leq k$, computed earlier and stored in $\Omega_k$ without any repeated time-consuming computation of the values of $w(y)$ from (\ref{eq:01}) i.e.
\begin{equation}
\label{eq:21}
w_i \xrightarrow{\lambda,P} (z_i,g_i ), 1 \leq i \leq k,\forall \lambda \in \Lambda, P \in \mathbb{P}.
\end{equation}

@@

The method of organizing the parallel computations considered above can be developed further by utilizing the information compatibility of the problems from the set $\mathbb{F}$ -- as has been noted in Section {\ref{sec:04}}, the values of characteristics $w^i$, $1 \leq i \leq k$, computed when solving any problem $F_1(\lambda,y(x))$ can be transformed to the values of any other problem $F_2(\lambda,y(x))$. This result allows proposing the following general scheme of parallel computations for simultaneous solving the problems from the set $\mathbb{F}$.

@@

\textbf{3. The execution of the global search iterations.} The execution of the global search iterations for each problem $F(\lambda,y(x))$ from the set $\mathbb{F}$ being solved is performed in parallel. The execution of every iteration on each computational node includes performing the following operations:
\begin{enumerate}
	\item The choice of $q$, $q \geq 1$ points $x^i$, $1\leq i\leq q$, of the next global search iteration is performed according to the rule (\ref{eq:20}) of the optimization method; the choice of the points is performed taking into account the available search information $\Omega_k$ and $A_k$; the number of generated points $q$, $q\geq 1$ is determined by the number of computational cores employed for solving the problem $F(\lambda,y(x))$;
	\item For each chosen one-dimensional point $x^i \in [0,1]$, $1 \leq i \leq q$, of current global search iteration, the multidimensional image $y^i \in D$, $1 \leq i \leq q$, is determined according to the mapping $y(x)$. Then, each computed image $y^i \in D$, $1 \leq i \leq q$, is transmitted to all employed computational nodes to avoid the multiple choice of the same points in the domain $D$ when solving the problems from the set $\mathbb{F}$;
	\item The values of characteristics $w(y)$ from (\ref{eq:01}) are computed at all points $y^i \in D$, $1 \leq i \leq q$; for each particular point $y^i \in D$, $1 \leq i \leq q$, these computations are performed in parallel using different computational cores; the computed values of characteristics $w(y)$ are transmitted to all employed computational nodes to include the obtained data into the search information $\Omega_k$ and $A_k$.
\end{enumerate}

@@

According to the presented computational scheme of parallel computations, each computational unit would contain the identical copies of the matrix of search information $\Omega_k$ from (\ref{eq:18}); the matrices $A_k$ from (\ref{eq:19}) would contain different sets of the global search points $x_i$, $1\leq i\leq k$, because of the use of different evolvents $y(x)$ for the dimensionality reduction and of different values of the scalar criterion and constraints $z_i$, $g_i$, $1\leq i\leq k$, corresponding to the problems $F(\lambda,y(x))$ from the set $\mathbb{F}$ being solved.

@@

The developed general computational scheme includes the concurrent parallel computational methods for the computational systems with the shared memory, the distributed parallel computations using the multiple mappings for the dimensionality reduction, and the multilevel nesting of the parallel computations for the high-performance computational systems. In general, the proposed computational scheme can provide the opportunity of an efficient application of the high-performance computational systems with a large number cores/processors for solving the complex global optimization problems. Besides, this general scheme can be used for the organization of the parallel computations for a wide range of algorithms for solving the time-consuming problems of decision making (in particular, for the information-statistical global optimization algorithms).

@@

\bibitem {x17_1}Gergel, V., Kozinov, E.: GPU-based computations in multicritical optimization . Communications in Computer Science and Information Science, \textbf{965}, 88--100 (2019). \doi{10.1007/978-3-030-05807-4\_8}
